---
output: github_document
header-includes: 
  \usepackage{amsmath}
---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-figures/"
)
```

## voila: Variational Inference for Langevin Equations
`voila` is an `R` package for the non-parametric estimation of Langevin
equations (also called Stochastic Differential Equations or SDE) from a 
densely-observed time series. The Langevin equation states that the dynamic
evolution of an observable X depends on a deterministic force, 
called the drift, and some random fluctuation, called the diffusion:

<img src=README-figures/langevin_annotated.png width=350 style="display: block; margin: auto;" >

where W represents a [Wiener process](https://en.wikipedia.org/wiki/Wiener_process).

`voila` permits to estimate the drift and diffusion terms by modelling them as
[gaussian processes](http://katbailey.github.io/post/gaussian-processes-for-dummies/)
(GPs). To cope with the computational complexity that calculating the posterior 
distribution of the GPs requires, the GPs are approximated using a small set of 
function points, the inducing variables.  These inducing variables are the
result of evaluating the drift  and diffusion terms at some _strategically_ 
located pseudo-inputs. The pseudo-inputs and the approximate posterior 
distributions are learnt using 
[variational inference](http://blog.evjang.com/2016/08/variational-bayes.html),
which minimizes the  Kullback-Leibler divergence between the true posterior 
distribution and the approximated one. 

The method is fully described in the paper:

> García, C.A., Otero, A., Félix, P., Presedo, J. & Márquez D.G., (2017). **Non-parametric Estimation of Stochastic Differential Equations with 
Sparse Gaussian Processes** _(under review)_,  [preprint](https://arxiv.org/abs/1704.04375).

## Installation
`voila` is not currently available on [CRAN](http://cran.r-project.org/) but it
may be installed from github using [devtools](https://github.com/hadley/devtools).
For the moment, `voila` requires the use of [yuima](https://yuima-project.com)
version 1.4.7, since the simulation method has changed in the latest version 
(this issue will be solved in the future). Hence, to install `voila`,
[yuima_1.4.7](https://cran.r-project.org/src/contrib/Archive/yuima/yuima_1.4.7.tar.gz) 
should be downloaded. The following chunk of code can be used for the download 
and the installation:

```R
url = 'https://cran.r-project.org/src/contrib/Archive/yuima/yuima_1.4.7.tar.gz'
download.file(url, 'yuima_1.4.7.tar.gz')
install.packages('yuima_1.4.7.tar.gz', repos = NULL)
# install all the remaining dependencies
install.packages(c('KernSmooth', 'doParallel', 'foreach', 'zoo',
                  'Rcpp', 'RcppArmadillo'))
library("devtools")
install_github("citiususc/voila", dependencies = FALSE)
```

## A quick-example
In this quick-example, `voila` is used to estimate the drift and diffusion terms  
from a single realization of an unidimensional
[Ornstein–Uhlenbeck](https://en.wikipedia.org/wiki/Ornstein%E2%80%93Uhlenbeck_process)
process.

```{r ornstein_example, cache=TRUE, fig.align='center'}
library("voila", quietly = TRUE, verbose = FALSE)
# simulate a Ornstein-Uhlenbeck time series using voila ---------------------
set.seed(1234)
samplingPeriod = 0.001
drift = "-x"
diffusion = "sqrt(1.5)"
x = simulate_sde(drift, diffusion, samplingPeriod = 0.001, tsLength = 20000)
plot.ts(x, ylab = "x(t)", xlab = "Time t", main = "Ornstein–Uhlenbeck process")
# Variational Inference  ----------------------------------------------------
## Some parameters for the algorithm
# the number of inducing points
noInducingPoints = 10 
# Our prior belief about the amplitude of the drift and diffusion functions
functionsUncertainty = 5 
# A small value to be added to the diagonal of the covariance matrix for
# stability purposes
epsilon = 1e-5
# The dimensionality of the data
inputDim = ncol(x)
# Since the data is 1D, we shall infer the equations for the 1st
# (and unique) dimension:
targetIndex = 1

# Select some initial pseudo-inputs. The positions of the pseudo-inputs
# are optimized during the inference
pseudoInputs = matrix(seq(min(x), max(x), len = noInducingPoints), ncol = 1)

# Create the kernels defining the behaviour of the gaussian processes. 
# Create a Rational Quadratic Kernel for the drift with some initial values for
# the hyperparameters. These hyperparameters will be optimized during the 
# inference process.
driftKer = sde_kernel("rq_kernel",
                       list('amplitude' = functionsUncertainty,
                            'alpha' = 1,
                            'lengthScales' = 1.5),
                       inputDim, epsilon)
# Create an Exponential kernel + constant term for the diffusion:
# k(x,x') = (maxAmplitude - expAmplitude) + expAmplitude * exp(- (x - x') ^ 2 / (2 * lengthScale))
# Voila uses a log-normal prior to ensure the positiveness of the diffusion. The 
# 'select_diffusion_parameters' function permits to select a proper amplitude for the
#  kernel from our prior belief about the amplitude of the diffusion function. It also
# selects a mean value for the log-normal distribution (denoted with v)
diffParams = select_diffusion_parameters(x, samplingPeriod, 
                                         priorOnSd = functionsUncertainty)
diffKer =  sde_kernel("exp_const_kernel",
                      list('maxAmplitude' = diffParams$kernelAmplitude,
                           'expAmplitude' = diffParams$kernelAmplitude * 1e-3,
                           'lengthScales' = 1.5),
                      inputDim, epsilon)

# Perform the variational inference (VI)
inference = sde_vi(targetIndex, x, samplingPeriod, pseudoInputs, 
                   driftKer, diffKer, diffParams$v)
# Analyze results  --------------------------------------------------------
# Check convergence
plot(inference$likelihoodLowerBound[-1], type = "o",
     main = "Likelihood Lower Bound", ylab = "Lower Bound", xlab = "Iteration")

# Get the estimations for the drift and diffusion
predictionSupport = matrix(seq(quantile(x,0.05), quantile(x,0.95), len = 100),
                           ncol = 1)
driftPred = predict(inference$drift, predictionSupport)
# the diffusion uses a log-normal gaussian process, so we must specify log = TRUE
diffPred = predict(inference$diff, predictionSupport, log = TRUE)

# Plot drift
realDrift = 
plot(driftPred, main = "voila's Drift Estimate",
     xlab = "x", ylab = "Drift f(x)")
lines(predictionSupport, eval(parse(text = drift), list(x = predictionSupport)),
      col = 2)
legend("topright", lty = 1, col = 1:2,
       legend = c("Estimate", "Real"), bty = "n")

# Plot diffusion
plot(diffPred, ylim = c(1, 2), main = "voila's Diffusion Estimate",
     xlab = "x", ylab = "Diffusion g(x)")
abline(h = eval(parse(text = diffusion), list(x = predictionSupport)) ^ 2,
       col = 2)
legend("topright", lty = 1, col = 1:2,
       legend = c("Estimate", "Real"), bty = "n")
```

## vignettes
A vignette is a long-form guide used for `R` packages. The `voila` vignettes
can be accessed after installing the package by typing:

```R
vignette(vignetteName, package = 'voila')
```

Currently, the following vignettes are available:
  
  * `'do_events'`: `voila` is applied for studying the Dansgaard-Oeschger (DO) 
  events. The DO events are fast climate changes that occurred during the last 
  glacial period.
  * `'multivariate_analysis'`: an example of how to use `voila` to study 
  a multi-dimensional time series.

## License
This project is licensed under the terms of the [GPL v3 license](LICENSE).

