---
title: "Real example: analysis of DO events"
author: "Constantino A. García"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{do_events}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
header-includes: 
  \usepackage{bm}
---
In this vignette, a paleoclimate record from the Greenland ice cores is studied.
Among other interesting phenomena, these records permit to identify the abrupt 
climate changes that occurred during the last glacial period, which are usually
referred to as  Dansgaard-Oeschger (DO) events. Specifically, a 
$\delta^{18}\text{O}$ record from the North Greenland Ice Core Project
([NGRIP](http://www.iceandclimate.nbi.ku.dk/data/)) is used. The 
$\delta^{18}\text{O}$ is a measure of the  ratio of the stable isotopes 
oxygen-18 and oxygen-16 and it is commonly used to estimate the temperature at 
the time that each small section of the ice core was formed. The data is already
preloaded with `voila`:

```{r plot_do, fig.align='center', message=FALSE}
library('voila')
plot(do_events, main = "DO events",
     xlab = "Time (Ky before present)", 
     ylab = expression(paste(delta,'18-O (permil)')))
```

For the sake of simplicity (and convergence speed), we shall use 10
pseudo-inputs and two rational quadratic kernels, although these selections are
not optimal (see reference [1] for further details). The following snippet performs
the inference after specifying some initial values for the pseudo-inputs and 
the kernels.

```{r inference, fig.align='center', message=FALSE, results = 'hide', cache=TRUE, eval=FALSE}
# sampling period in Ky before present
samplingPeriod = deltat(do_events) 
# voila uses data in matrix format 
do_events = matrix(do_events, ncol = 1)
# Spread initial pseudo-inputs along the support of do_events
noInducingPoints = 10
pseudoInputs = matrix(seq(min(do_events), max(do_events), len = noInducingPoints),
                      ncol = 1)
# small value to regularize covariance matrices
epsilon = 1e-5
# Create the kernels defining the behaviour of the gaussian processes. 
# We shall create two Rational Quadratic Kernel for both the drift and 
# diffusion with some initial values for the hyperparameters. These 
# hyperparameters will be optimized during the inference process.
driftKer = sde_kernel("rq_kernel",
                       list('amplitude' = 30, # huge amplitude: huge uncertainty
                            'alpha' = 2,
                            'lengthScales' = 2),
                      ncol(do_events), epsilon)
# Voila uses a log-normal prior to ensure the positiveness of the diffusion. The 
# 'select_diffusion_parameters' function permits to select a proper amplitude for the
# kernel from our prior belief about the amplitude of the diffusion function. It also
# selects a mean value for the log-normal distribution (denoted with v)
diffParams = select_diffusion_parameters(do_events, samplingPeriod, 
                                         priorOnSd = 30)
diffKer =  sde_kernel("rq_kernel",
                      list('amplitude' = diffParams$kernelAmplitude,
                           'alpha' = 1,
                           'lengthScales' = 2),
                      ncol(do_events), epsilon)

# Perform the variational inference (VI)
inference = sde_vi(1, do_events, samplingPeriod, pseudoInputs, 
                   driftKer, diffKer, diffParams$v,
                   maxIterations = 1000)

```

```{r loadData, echo=FALSE}
# voila ships with the results of the previous snippet to avoid slow
# installations
inference = readRDS("../inst/extdata/oxygen_estimates.RDS")

```

The following snippet plots the resulting estimates of the drift and diffusion
functions.

```{r plot_results, fig.align='center'}
# Get the estimations for the drift and diffusion
predictionSupport = matrix(seq(quantile(do_events,0.025), 
                               quantile(do_events,0.975), len = 100),
                           ncol = 1)
driftPred = predict(inference$drift, predictionSupport)
# the diffusion uses a log-normal gaussian process, so we must specify log = TRUE
diffPred = predict(inference$diff, predictionSupport, log = TRUE)

# Plot drift
plot(driftPred, main = "voila's Drift Estimate",
     xlab = expression(paste("x = ", delta,'18-O (permil)')),
     ylab = "Drift f(x)")
abline(h = 0, lty = 2, col = 2)
# Plot diffusion
plot(diffPred, main = "voila's Diffusion Estimate",
     xlab = expression(paste("x = ", delta,'18-O (permil)')),
     ylab = "Diffusion g(x)")
```

Note that the drift function has two stable points (where the drift function
crosses zero with negative slope) for x = -43.5 and x = -40, corresponding to
the cold stable state and the hot metastable state, respectively. The diffusion
function supports the use of a state-dependent diffusion 
rather than the widely-used constant term. Further discussion about the diffusion
function may be found at reference [1].


### References
This example is discussed in detail in the paper:

> [1] García, C.A., Otero, A., Félix, P., Presedo, J. & Márquez D.G., (2017). **Non-parametric Estimation of Stochastic Differential Equations with 
Sparse Gaussian Processes** _(under review)_,  [preprint](https://arxiv.org/abs/1704.04375).

