---
title: "Multivariate Langevin Equations"
author: "Constantino A. García"
#output: rmarkdown::html_vignette
output: html_document
header-includes: 
  \usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{multivariate_equations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

Let's consider the multivariate Langevin equation for a set of N random variables
$\mathbf{X} = [X^{(1)}, X^{(2)}, \cdots, X^{(N)}]$:

$$dX^{(i)}_t = f_i(\mathbf{X}_t)d{t}  + \sum_j \sqrt{g_{ij}(\mathbf{X}_t)} d{W^{(j)}_t}.$$ 

Our aim is to compute all the unknown functions $f_i$ and $g_{ij}$ from a single
observation of the (multivariate) time series $\mathbf{X}_t$. Currently, the general case with a full diffusion matrix $\mathbf{G}$ cannot be solved with `voila`.
However,  if the matrix is diagonal, we can still apply the estimation method 
implemented in the package.

### An example
First of all, we shall simulate a 2D time series. The equations do not intend 
to have a physical meaning, but the are inspired by the harmonic oscillator 
equation.

```{r simulation, message=FALSE, fig.align='center', eval=TRUE}
suppressMessages(library("voila"))
# simulate a multivariate time series using voila ---------------------
set.seed(100)
driftExpression = c("x2", "-2 ^ 2 * x1")
diffExpression = matrix(c("sqrt(1.25)", 0,
                          0, "sqrt(0.75 + 0.25 * x1 ^ 2 + 0.5 * x2 ^ 2)"),
                        2, 2)
samplingPeriod = 0.001
x  = simulate_sde(driftExpression,
             diffExpression,
             samplingPeriod, 12000,
             stateVariable = c("x1", "x2"))
xTime = seq(0, length.out = nrow(x), by = samplingPeriod)
par(mfrow = c(2,1), mar = c(4,4,0.25,0.25))
plot(xTime, x[,1], type = 'l', ylab = 'x1', xlab = '')
plot(xTime, x[,2], type = 'l', ylab = 'x2', xlab = 'time')
```

Now, we try to estimate the drift and diffusion terms corresponding to
$X^{(2)}$:

```{r inference,eval=TRUE,cache=TRUE, eval = FALSE}
## Some parameters for the algorithm
# the number of inducing points
noInducingPoints = 10 
# Our prior belief about the amplitude of the drift and diffusion functions
uncertainty = 5 
# A small value to be added to the diagonal of the covariance matrix for
# stability purposes
epsilon = 1e-5
# The dimensionality of the data
inputDim = ncol(x)
# we shall infer the equations for the 2nd dimension
targetIndex = 2

# Selection of the pseudo-inputs. We shall use a quick an dirty way of getting
# some initial pseudo-inputs: we fit a multivariate normal distribution to the
# data and then sample the distribution (although the data it is not normally 
# distributed!!).
library('mvtnorm')
xm = rmvnorm(10, colMeans(x), cov(x))

# Create the kernels
diffParams = select_diffusion_parameters(x, samplingPeriod, priorOnSd = uncertainty,
                                         responseVariableIndex = targetIndex)
driftKer = sde_kernel('exp_kernel', list('amplitude' = uncertainty,
                                         'lengthScales' = c(2,2)),
                      inputDim, epsilon)

diffKer = sde_kernel('exp_kernel', list('amplitude' = diffParams$kernelAmplitude,
                                          'lengthScales' = c(0.5,0.5)),
                     inputDim, epsilon)

# Variational inference. The number of iterations could be increased, but we want
# a quick result ;) 
inference = sde_vi(timeSeriesIndex = targetIndex, x, samplingPeriod, xm,
                   driftKer, diffKer, diffParams$v)
```

```{r loadData, echo = FALSE}
# voila ships with the results of the previous snippet to avoid slow
# installations
targetIndex = 2
inference = readRDS("../inst/extdata/multivariate_inference.RDS")
```


Finally we plot the resulting estimates to compare them with the true
drift and diffusion functions. 

```{r plots,fig.align='center', eval = TRUE}
library('plot3D')
# The new input data to generate the predictions
x1 = seq(quantile(x[,1], 0.05), quantile(x[,1], 0.95), len = 100)
x2 = seq(quantile(x[,2], 0.05), quantile(x[,2], 0.95), len = 100)

## To generate the 3D-plot we have to generate a grid of function points
# Drift function evaluation
zDrift = outer(x1, x2, function(x1, x2) {
  eval(parse(text = driftExpression[[targetIndex]]),
       list(x1 = x1, x2 = x2))

})
zDriftEst = outer(x1, x2, function(x1, x2) predict(inference$drift,
                                                  matrix(c(x1,x2), ncol = 2),
                                                  lognormal = FALSE)$mean
)
# Diffusion function evaluation
zDiff = outer(x1, x2, function(x1, x2) {
  eval(parse(text = diffExpression[targetIndex,targetIndex]),
       list(x1 = x1, x2 = x2)) ^ 2 + rnorm(length(x1), sd = 0.001)

})
zDiffEst = outer(x1, x2, function(x1, x2) predict(inference$diff,
                                                  matrix(c(x1,x2), ncol = 2),
                                                  lognormal = TRUE)$mean
)
# Drift plot
par(mfrow = c(2,2), mar = c(0.1,0.1,2,0.1))
col = seq(min(c(zDrift, zDriftEst)), max(c(zDrift, zDriftEst)), len = 25)
persp3D(x = x1,y = x2, zDrift, phi = 35, #ticktype = 'detailed',
        zlim = range(c(zDriftEst, zDrift)), breaks = col, colkey = FALSE,
        main = 'Real Drift', xlab = 'x1', ylab = 'x2',
        zlab = 'f_2(x1, x2)')
persp3D(x = x1,y = x2, zDriftEst, phi = 35, #ticktype = 'detailed',
        zlim = range(c(zDriftEst, zDrift)), breaks = col, colkey = FALSE,
        main = 'Drift Estimate', xlab = 'x1', ylab = 'x2',
        zlab = 'f_2(x1, x2)')
# Diff plot
col = seq(min(c(zDiff, zDiffEst)), max(c(zDiff, zDiffEst)), len = 25)
persp3D(x = x1,y = x2, zDiff, phi = 35, #ticktype = 'detailed',
        zlim = range(c(zDiff, zDiffEst)), breaks = col, colkey = FALSE,
        main = 'Real Diffusion', xlab = 'x1', ylab = 'x2',
        zlab = 'g_22(x1, x2)')
persp3D(x = x1,y = x2, zDiffEst, phi = 35, #ticktype = 'detailed',
        zlim = range(c(zDiff, zDiffEst)), breaks = col, colkey = FALSE,
        main = 'Diffusion Estimate', xlab = 'x1', ylab = 'x2', 
        zlab = 'g_22(x1, x2)')
```
